---
title: "Homework 5"
author: "Zara Mecklai"
date: April 24, 2026
format: pdf
engine: knitr
---

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, fixest)

# HCRIS data 
final.hcris.data <- read_csv('../data/output/HCRIS_final.csv')

# State crosswalk 
state.crosswalk <- tibble(
  state = c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA",
            "HI","ID","IL","IN","IA","KS","KY","LA","ME","MD",
            "MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ",
            "NM","NY","NC","ND","OH","OK","OR","PA","RI","SC",
            "SD","TN","TX","UT","VT","VA","WA","WV","WI","WY","DC"),
  State = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado",
            "Connecticut","Delaware","Florida","Georgia","Hawaii","Idaho",
            "Illinois","Indiana","Iowa","Kansas","Kentucky","Louisiana",
            "Maine","Maryland","Massachusetts","Michigan","Minnesota",
            "Mississippi","Missouri","Montana","Nebraska","Nevada",
            "New Hampshire","New Jersey","New Mexico","New York",
            "North Carolina","North Dakota","Ohio","Oklahoma","Oregon",
            "Pennsylvania","Rhode Island","South Carolina","South Dakota",
            "Tennessee","Texas","Utah","Vermont","Virginia","Washington",
            "West Virginia","Wisconsin","Wyoming","District of Columbia")
)

# KFF data 
kff.dat <- read_csv('kff_data.csv', skip=2)
kff.final <- kff.dat %>%
  rename(State = Location,
         Expansion.Status = `Status of Medicaid Expansion Decision`,
         date_adopted = `Expansion Implementation Date`) %>%
  filter(State %in% state.crosswalk$State) %>%
  mutate(expanded = (Expansion.Status == "Adopted"),
         date_adopted = mdy(date_adopted)) %>%
  select(State, expanded, date_adopted)

# Merge 
final.data <- final.hcris.data %>%
  left_join(state.crosswalk, by="state") %>%
  left_join(kff.final, by="State") %>%
  mutate(expand_year = year(date_adopted),
         expand = (year >= expand_year & !is.na(expand_year))) %>%
  rename(expand_ever = expanded) %>%
  mutate(uncomp_care_final = case_when(
    !is.na(uncomp_care) ~ uncomp_care,
    TRUE ~ tot_discounts + replace_na(bad_debt, 0)
  )) %>%
  filter(uncomp_care_final > 0 | is.na(expand_ever)) %>%        
  filter(is.na(expand_year) | expand_year == 2014) %>%          
  mutate(post = (year >= 2014),
         treat = as.integer(post) + as.integer(expand_ever))
```

Question 1: Mean uncompensated care by year (in millions) 
```{r}
final.data <- final.hcris.data %>%
  left_join(state.crosswalk, by="state") %>%
  left_join(kff.final, by="State") %>%
  mutate(expand_year = year(date_adopted),
         expand = (year >= expand_year & !is.na(expand_year))) %>%
  rename(expand_ever = expanded) %>%
  mutate(uncomp_care_final = bad_debt + tot_uncomp_care_charges - tot_uncomp_care_partial_pmts) %>%
  filter(uncomp_care_final > 0 | is.na(expand_ever)) %>%
  filter(is.na(expand_year) | expand_year == 2014) %>%
  mutate(post = (year >= 2014),
         treat = as.integer(post) + as.integer(expand_ever))

q1.table <- final.data %>%
  ungroup() %>%
  filter(year >= 2011, year <= 2018) %>%
  group_by(year) %>%
  summarize(
    mean_uncomp_care = mean(uncomp_care_final, na.rm=TRUE) / 1000000,
    n_hospitals = n_distinct(provider_number)
  ) %>%
  ungroup()

knitr::kable(q1.table,
             col.names = c("Year", "Mean Uncompensated Care (Millions)", "Hospitals"),
             digits = 2,
             caption = "Mean Hospital Uncompensated Care by Year (2011-2018)")
```
Interpretation: Uncompensated care has stayed relatively constant. 


Question 2: Plot mean uncompensated care by expansion status 
```{r}
q2.data <- final.data %>%
  ungroup() %>%
  filter(year >= 2010, year <= 2018) %>%
  filter(!is.na(expand_ever)) %>%   # remove NA hospitals
  mutate(expand_group = ifelse(expand_ever == FALSE, "Never Expanded", "Expanded 2014")) %>%
  group_by(year, expand_group) %>%
  summarize(mean_uncomp_care = mean(uncomp_care_final, na.rm=TRUE) / 1000000, .groups="drop")

ggplot(q2.data, aes(x=year, y=mean_uncomp_care, color=expand_group)) +
  geom_line(size=1) +
  geom_point(size=2) +
  geom_vline(xintercept=2014, linetype="dashed", color="grey50") +
  labs(title = "Mean Hospital Uncompensated Care by Medicaid Expansion Status",
       x = "Year",
       y = "Mean Uncompensated Care (Millions $)",
       color = "Expansion Status",
       caption = "Dashed line indicates 2014 Medicaid expansion") +
  theme_bw()
```
Interpretation: The graph does suggest a treatment effect as the expanded uncompensated care diverges from the never expanded uncompensated care. 

Question 3: 2x2 DD table (2012 pre, 2015 post) 
```{r}
q3.data <- final.data %>%
  ungroup() %>%
  filter(year %in% c(2012, 2015)) %>%
  filter(!is.na(expand_ever)) %>%
  mutate(post = ifelse(year == 2015, "2015 (Post)", "2012 (Pre)"),
         group = ifelse(expand_ever == TRUE, "Expansion States", "Non-Expansion States")) %>%
  group_by(group, post) %>%
  summarize(mean_uncomp_care = mean(uncomp_care_final, na.rm=TRUE) / 1000000, .groups="drop") %>%
  pivot_wider(names_from = post, values_from = mean_uncomp_care) %>%
  rename(Group = group)

knitr::kable(q3.data,
             digits = 2,
             col.names = c("Group", "2012 (Pre)", "2015 (Post)"),
             caption = "2x2 DD Table of Mean Uncompensated Care (Millions $)")
```


Question 4: Trend Discussion
The non-expansion states saw uncompensated rates rise, likely due to general healthcare cost inflation and post-ACA coverage changes. The expansion group decreased, consistent with Medicaid picking up previously uninsured patients. Medicaid expansion would be expected to reduce uncompensated care because newly insured patients who previously couldn't pay now have coverage, reducing the burden on hospitals.

Question 5: Standard DD regression
```{r}
q5.reg <- lm(uncomp_care_final/1000000 ~ expand + year + expand_ever,
             data = final.data %>% ungroup() %>% filter(year >= 2010, year <= 2018))

summary(q5.reg)
```

Question 6: DD with hospital and year fixed effects 
```{r}
q6.reg <- feols(uncomp_care_final/1000000 ~ expand | provider_number + year,
                cluster = ~state,
                data = final.data %>% ungroup() %>% filter(year >= 2010, year <= 2018))

summary(q6.reg)
```
Interpretation: The point estimate decreases (more negative). 

Question 7: DD with all states 
```{r}
q7.data <- final.hcris.data %>%
  left_join(state.crosswalk, by="state") %>%
  left_join(kff.final, by="State") %>%
  mutate(expand_year = year(date_adopted),
         expand = (year >= expand_year & !is.na(expand_year))) %>%
  rename(expand_ever = expanded) %>%
  mutate(uncomp_care_final = bad_debt + tot_uncomp_care_charges - tot_uncomp_care_partial_pmts) %>%
  filter(uncomp_care_final > 0 | is.na(expand_ever)) %>%
  filter(year >= 2010, year <= 2018)

q7.reg <- feols(uncomp_care_final/1000000 ~ expand | provider_number + year,
                cluster = ~state,
                data = q7.data %>% ungroup())

summary(q7.reg)
```
Interpretation: Now the data is capturing the effect for states that expanded in 2015, 2016, 2017, and 2018. These late expanders may have expanded under different conditions, and their "post" period is much shorter within the 2010-2018 window. This dilutes the estimated effect.

Question 8: Event study - 2014 expanders and never-expanders only 
```{r}
q8.data <- final.data %>%
  ungroup() %>%
  filter(year >= 2010, year <= 2018) %>%
  mutate(relative_year = ifelse(expand_ever == TRUE, year - 2014, -1),
         relative_year = factor(relative_year))

q8.reg <- feols(uncomp_care_final/1000000 ~ i(relative_year, expand_ever, ref="-1") |
                provider_number + year,
                cluster = ~state,
                data = q8.data)

iplot(q8.reg,
      main = "Event Study: Effect of Medicaid Expansion on Uncompensated Care",
      xlab = "Years Relative to Expansion (2014)",
      ylab = "Effect on Uncompensated Care (Millions $)")
```

Question 9: Event study - all states with binned event time 
```{r}
q9.data <- q7.data %>%
  ungroup() %>%
  mutate(relative_year = ifelse(expand_ever == TRUE, year - expand_year, -1),
         relative_year = case_when(
           relative_year < -4 ~ -4,
           relative_year > 4  ~  4,
           TRUE               ~ relative_year
         ),
         relative_year = factor(relative_year))

q9.reg <- feols(uncomp_care_final/1000000 ~ i(relative_year, expand_ever, ref="-1") |
                provider_number + year,
                cluster = ~state,
                data = q9.data)

iplot(q9.reg,
      main = "Event Study: Effect of Medicaid Expansion on Uncompensated Care (All States)",
      xlab = "Years Relative to Expansion",
      ylab = "Effect on Uncompensated Care (Millions $)")
```

Question 10: Summary of Findings (Questions 5-9)
Medicaid expansion is associated with a reduction in hospital uncompensated care. Q8 and Q9 provide support for parallel trends as the pre-period coefficients are small and close to zero, though not perfectly flat. This suggests the parallel trends assumption is approximately satisfied for 2014 expanders versus never-expanders. One limitation of the standard TWFE estimator in staggered designs is that it uses already-treated units as controls for not-yet-treated units, creating "forbidden comparisons" that can bias estimates toward zero. This suggests the true effect of Medicaid expansion on uncompensated care is better captured by the cleaner 2014-versus-never-expander comparisons in Q6 and Q8.

